---
title: "Statistical Analysis Using R Unit-1"
author: "MOHAMMAD WASIQ"
date: "9/4/2020"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
# Introduction to R language

## What is R ?

**R** is a language and environment for statistical computing . It has following features:

* It is free and can be downloaded from intenet: http://cran.r-project.org. It requires only 80 Mb spaces.

*It is a functional language.

* It has very powerful graphic capabilities.

* It has capabilities to interface with other important packages for data analysis and graphics.

## Creating a vector in R---'c()'

Let us define a vector $x$ as 
```{r vect1}
x=c(5,4,3,1,4) # define x vector
x
length(x)
```

Note that R commands are written in a chunk which begins with three back ticks and ends with three back ticks. The command ```{r vect1}``` specifies that chunk is for R language and it is labeled by 'vect1'. All the R commands are to be written inside the chunk.

## Arithmetic Operations on vectors

There are five basic operators in R, namely '+, -, *, / ^' , which   are meant for addition , negation , multiplication , dividatiuon , exponentiation respectively. Besides these basic operators, a large  number of function are 
also available for arithmetic computations. Let us create a new chumk:

```{r arithmeticopvect1}
sum(x) # sum of elements 
sum(x^2) # sum of squares 
sqrt(x) # square root of x 
log(x) # log of x with base e
log(x, base=10)# log of x with 10
log(x, base=5) # log of x with 5
1/x # inverse of x
x^-1 # same as above
sin(x) 
cos(x)
tan(x)
mean(x)
var(x) # variance of x
sd(x) # standard deviation of x
cv=function(x) sd(x)/mean(x)*100
dump("cv", file="cv.txt")
source("cv.txt")
cv(x)
```
## Concept of recycling rule

The basic rele is that smaller vector recycles upto the length of the larger vector. Let us see the commands in the following chunk;
```{r rec1}
x=c(5,4,3,1,4)
x+1 # 1 will recycle up to the length of x andd add
2*x
log(x+1)
sin(2*x)
```

## The function repeat---'rep()'
```{r rep1}
rep(c(1,2),c(3,3)) # repeat 1 3 times and 2 also 3 times
rep(c(1,2),each=4)
tasbihefatimi=rep(c("SubhanAllah","Alhamadulliah","Allahuakbar"),c(33,33,34))
tasbihefatimi
rep(c("R1","R2","R3"),c(3,2,4))
```
## Logical operators---TRUE(1),FALSE(0)
```
>, <, <=,>=, == exactly equal to, != not equal to
```
```{r logic1}
x=c(2,5,3,1,8)
x>3
sum(x>3)
(x[x>2])
mean(x[x>2])
mean(x[x<3])
mean(x<3)
#logical expression | or, & , and
x>2 & x<4
x[x>2 & x<4]
```
## Missing values--- NA is.na()

```{r missing1}
x=c(2,5,3,NA,6)
is.na(x)
mean(x)
mean(x,na.rm=TRUE)
!is.na(x)
mean(x[!is.na(x)])
## Not a number, NaN, 0/0, Inf-Inf, Inf/Inf
x=c(0,5,3)
x/x
is.na(x/x)
x=c(2,5,8)
x
names(x)=c("Zainul","Altamash","Wasiq")
x
barplot(x)
```

**paste()**
The function 'paste()' behave like 'c()' but it always returns a character vector.

**letters** and **LETTERS**

```{r let1}
letters
LETTERS
letters[c(3,6)]
LETTERS[1:6]
month.abb
month.abb[c(3,9)]
month.abb[3:9]
month.name
```

**substrin**

```{r sub1}
substr(month.name,1,4)
```

## Factor vector---'factor()', 'ordered()'
Factor is like the 

```{r fact1}
grp=c("control","treatment","control","treatment")
grp
grp=factor(grp)
as.integer(grp)
# factor are efficient ways of storing data
as.integer(grp)
levels(grp)
```

## Matrices and arrays--- 'matrix()' , 'arrays()'
The function 'matrix()' is used to create a matrix.
```{r mat1}
m=matrix(1:6,nrow=2,ncol=3)
m
dim(m)
dimnames(m)=list(LETTERS[1:2],letters[1:3])
m
colnames(m)=c("Samad","Radha","Umar")
m
rownames(m)=c("Moradabad","Aligarh")
m
matrix(1:6,ncol=3,byrow=TRUE)
# Extraction of a matrix---[]
m[1,3]
m[,c(1,2)]
m[c(1,2),3]
## arrays()
array(1:8,c(2,2,2))
# naming arrays
dim(array)
```

## Data frame---'data.farme()'
These are objects, a kind of generalization of matrices. Generalization in the sense that columns can be of different types from each other. This flexibility is not with matrices. Use 'data.frame()' to construct a data from vector:

```{r df1}
colors=c("red","yellow","blue")
numbers=c(1,2,3)
colors_numbers=data.frame(colors, numbers)
colors_numbers
colors=c("red","yellow","blue")
numbers=c(1,2,3)
more_numbers=c(5,4,7)
colors_numbers=data.frame(colors,numbers,more_numbers)
colors_numbers
colors=c("red","yellow","blue")
numbers=c(1,2,3)
more_numbers=c(5,4,7)
colors_numbers=data.frame(Colors=colors,Numbers=numbers,MoreNUMBERS=more_numbers)
colors_numbers
# change row names
row.names(colors_numbers)=paste0("Row",1:3)
colors_numbers
# extraction from data.frame
colors_numbers[c(1,3),]
DF2=colors_numbers[,c(1,3)]
DF2
```


## Built in graphic function---'hist()', 'plot()', 'curve()'
**  Histogram **
```{r hist}
x=c(12,15,13,20,14,16,10,10,8,15)
hist(x)
hist(x,prob=TRUE)
## Scatter plot
x=seq(1,10)
y=x^2-10*x
plot(y~x)
## Curve
curve(x^2,from=-2,to=2)
curve(expr=sin,from=0,to=6*pi)
curve(dnorm(x),from=-3,to=3)
## Suppose you want to plot a function which is not defined in R.
# Plot f1(x)=2x+3x^2, -2 to 3
f1=function(x) 2*x+3*x^2
curve(f1,from=-2,to=3)
curve(f1,from=-3,to=3)
curve(f1,from=-3,to=3,lwd=2)
curve(f1,from=-3,to=3,lwd=2,col="red")
curve(f1,from=-3,to=3,lwd=3,col="blue")
curve(f1,from=-3,to=3,lwd=5,col="green")
```

## Histogram, Frequency Curve, Polygon, Ogive for a data
```{r ogive1}
# weight of 30 students
weight=c(68,53,69.5,55,71,63,76.5,65.5,69,75,76,57,70.5,71.5,56,81.5,69,59,67.5,61,68,59.5,56.5,73,61,72.5,71.5,59.5,74.5,63)
dump("weight","weight.dat")
length(weight)
# Create Histogram
h1=hist(weight)
h1
h1$breaks
# Frequency polygon on histogram itself
lines(h1$mids,h1$counts,type="o")
plot(h1$mids,cumsum(h1$counts),type="o")
stem(weight)
# Create a pie chart for the same data
pie(h1$count,label=h1$mids)
```

##Graphics with R
There are two type of graphic functions:

1. High level graphic functions which always create a new plot. Examples are; plot(),hist(), barplot(),
dotchart(), pie(), curve().

2. Low level graphic functions: They add on existing plot. For example, lines(), points(), mtext(),
main(), sub().

**Example**: Bar charts and dot charts The data VADeaths a data set in R contains death rates (number of
deaths per 1000 population per year) in various subpopulations within state of Viginia in 1940. This may
be plotted as barchart.
```{r vaeath1}
VADeaths
barplot(VADeaths, beside = TRUE, legend=TRUE,ylim=c(0,90),
ylab = "Deaths per 1000", main="Death rates in Virginia")
```

# Dot Chart---'dotchart()'
Position of dot of corresponding number.
```{r dotchart1}
dotchart(VADeaths, xlim=c(0,75), xlab="Deaths per 1000", main="Death rate in Virginia")
```

## Pie chart
Pie chart displays a vector of numbers by breaking up a circular disk into pieces whose angle (and hence
area) is proportional to each number. For example, the letter grades assigned to each class might arise in
the proportions shown in the following figure:
```{r pie1}
groupsizes=c(18,30,32,10,10)
labels=c("A","B","C","D","F")
pie(groupsizes, labels, col=c("grey40", "white","grey","black", "grey90"))
```

## Histogram--—hist()
**Example for histogram**
```{r hist2}
# simulate 5000 observations from N(0,1)
# make a histogram for the simulated data
x=rnorm(n=1000,mean=0,sd=1)
x
mean(x)
sd(x)
hist(x,prob=TRUE)
curve(dnorm(x,mean=0,sd=1),add=TRUE,lwd=2) #to add on histogram
# An example with exponential
xe=rexp(1000,rate=0.06)
mean(xe)
hist(xe, prob=TRUE)
curve(dexp(x,rate=0.06),add=TRUE)
curve(dexp(x,rate=0.10),add=TRUE)
```

```{r weib1}
# Simulate 1000 random values from Weibull(shape=1.2,scale=16)
xw=rweibull(n=1000,shape=1.5, scale=80)
hist(xw,prob=TRUE,ylim=c(0,.01))
curve(dweibull(x,shape=1.5,scale=80),add=TRUE)
```

## Details of histogram
```{r histod1}
x=rnorm(1000)
hist(x,prob=TRUE)
curve(dnorm(x),add=TRUE)
```
## Box plot---boxplot()
```{r weibullbox}
y=rweibull(1000,shape=1.5,scale=50)
hist(y,prob=TRUE)
boxplot(y,horizontal=TRUE)
median(y)
quantile(y,prob=c(0.25,.50,.75))
min(y)
max(y)
IQR(y)
LowerWhisker=max(min(y),20.449-1.5*IQR(y))
LowerWhisker
UpperWhisker=min(max(y),60.734+1.5*IQR(y))
UpperWhisker
```

## Box plot of 'Iris' data
```{r boxiris}
names(iris)
head(iris)
tail(iris)
str(iris)
boxplot(Sepal.Length~Species,data=iris)
```

## Scatter plot---'plot()'
```{r scatter1}
x= rnorm(100) # assigns 100 random normal observations to x
y=rpois(100,30)
mean(y)
mean(x)
sd(x)
plot(x,y,main="Poisson versus Normal")
plot(x,y,pch=16,type="points")
plot(x,y,type="l") #lines
plot(sort(x),sort(y),type="l")
```

## Plotting From a data frame
```{r plotDF}
data(Orange)
names(Orange)
head(Orange)
tail(Orange)
str(Orange)
Orange$Tree #extract column of Tree
unique(as.character(Orange$Tree))
plot(circumference~age, data = Orange)
plot(circumference~age, data = Orange, pch = as.character(Tree),
cex = 0.75)
```

## QQ plot
```{r qql}
z<-rnorm(10000,mean=0,sd=1)
quantile(z,prob=c(0.25,.50,.75))
par(mfrow = c(1,4))
X=rnorm(1000)
A=rnorm(1000)
qqplot(X,A,main="A and X are the same")
B=rnorm(1000,mean=3,sd=2)
qqplot(X,B,main="B is rescaled X")
C=rt(1000,df=2)
qqplot(X,C,main="C has heavier tails")
par(mfrow=c(1,1))
curve(dt(x,df=100),from=-5,to=5)
curve(dnorm(x),add=TRUE,lty=2)
```

# Low level graphic functions
## Adding to plots

* points(x, y, ...) # adds point
* lines(x, y, ...) # adds line segments
* text(x, y, labels, ...) # adds text into the graph
* abline(a, b, ...) # adds the line $y = a + bx$
* abline(h = y, ...) # adds a horizontal line
* abline(v = x, ...) # adds a vertical line
* polygon(x, y, ...) # adds a closed and possibly filled polygon
* segments(x0, y0, x1, y1, ...) # draws line segments
* arrows(x0, y0, x1, y1, ...) # draws arrows
* symbols(x, y, ...) # draws circles, squares, thermometers, etc.
  legend(x, y, legend, ...) # draws a *legend

# Linear Models--- 'lm()'
```{r lm1}
y=c(2,4,5,4,10)
x=c(4,5,10,12,10)
plot(y~x)
# y=beta0+beta1*x+error
M1=lm(y~x)
coef(M1)
abline(lm(y~x))
abline(M1,lwd=2,lty=2,col="red")
```
## Example Orange data
```{r circumfrence1}
plot(circumference~age, pch=as.numeric(as.character(Tree)),data=Orange)
abline(lm(circumference~age,data=Orange,subset=Tree=="1"))
abline(lm(circumference~age,data=Orange,subset=Tree=="2"),lty=2,lwd=2)
abline(lm(circumference~age,data=Orange,subset=Tree=="3"),lty=3)
abline(lm(circumference~age,data=Orange,subset=Tree=="4"),lty=4)
abline(lm(circumference~age,data=Orange,subset=Tree=="5"),lty=5)
legend("bottomright",legend = paste("Tree",1:5),lty=1:5,lwd=c(1,1,2,1,1))
```
```{r figAnn1}
par(mar=c(5,5,5,5)+0.1)
plot(c(1,9),c(0,50),type='n' ,xlab="", ylab="")
text(6, 40, "Plot region")
points(6, 20)
text(6, 20, "(6, 20)", adj = c(0.5, 2))
mtext(paste("Margin", 1:4), side = 1:4, line = 3)
mtext(paste("Line", 0:4), side = 1, line = 0:4, at = 3, cex = 0.6)
mtext(paste("Line", 0:4), side = 2, line = 0:4, at = 15, cex = 0.6)
mtext(paste("Line", 0:4), side = 3, line = 0:4, at = 3, cex = 0.6)
mtext(paste("Line", 0:4), side = 4, line = 0:4, at = 15, cex = 0.6)
```

## Axis and Tricks
```{r axix}
hist(log(1000*islands, 10), axes = FALSE, xlab = "Area (in sq. miles)",
main = "Areas of the World’s Largest Islands")
box()
axis(side = 1, at = 4:7, labels = c("10,000", "100,000", "1,000,000",
"10,000,000"))
axis(side = 2)
```

## Boxplot with some additional features
motor
## Brand 1 Brand 2 Brand 3 Brand 4 Brand 5
## 1 13.1 16.3 13.7 15.7 13.5
## 2 15.0 15.7 13.9 13.7 13.4
## 3 14.0 17.2 12.4 14.4 13.2
## 4 14.4 14.9 13.8 16.0 12.7
## 5 14.0 14.4 14.9 13.9 13.4
## 6 11.6 17.2 13.3 14.7 12.3
```{r box12}
data("motor")
boxplot(motor)
boxplot(motor, ylab = "Vibration", axes = FALSE)
title("Motor Vibration Data")
mtext("The Effects of Different Bearings", side = 3, line = 0)
box()
axis(side=2, las=2)
axis(side = 1, at = 1:5, label = names(motor))
```

